Introductory statistics with `intRo`

### 2016-02-19 jcgs-paper-2016/paper.Rmd
No more texttt

The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student. This additional time commitment may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors tend to use a tool like \texttt{R} in their own work. Still, some compromises must be made when using software like \texttt{JMP}, including dealing with software licenses and unsupported desktop platforms.

@@

Multiple software packages have recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic \texttt{R} programming and data analysis techniques. \texttt{Swirl} [@swirl] is a similar interactive tool to make learning \texttt{R} more fun by learning it within \texttt{R} itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach \texttt{R} programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

Upon the release of \texttt{RStudio}'s \texttt{Shiny} [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call \texttt{intRo}, available at http://www.intro-stats.com. \texttt{intRo} offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, \texttt{intRo} stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text.

# What is \texttt{intRo}?

\texttt{intRo} is a web-based tool to accompany an introductory statistics class. It is meant to assist in the learning of statistics rather than as a stand-alone delivery of statistics education, with the intention of being used in conjunction with a guided class. An accompanying R package, titled \texttt{intRo} and available on GitHub, assists in the downloading, running, and deploying of \texttt{intRo} instances.

Three fundamental philosophies that guided the creation of \texttt{intRo}. In particular, \texttt{intRo} is \emph{easy} to use and can be an \emph{exciting} part of learning statistics. Additionally, \texttt{intRo} is an \emph{extensible} tool, allowing for a user of \texttt{intRo} to tailor the tool for his or her own classroom needs.

In the development of \texttt{intRo}, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in \texttt{intRo}. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{images/user_experience.pdf}
\caption{A typical user experience of generating a result in \texttt{intRo}. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.}
\label{fig:user_experience}
\end{figure}

Beyond being simple, \texttt{intRo} is also consistent. The tool is organized around specific tasks a student may perform in the process of a data analysis, called modules. To the user, a module is simply a page of statistics functionality that maintains a consistent layout, helping the user to become familiar with the location of the options, the results, and the code. Figure \ref{fig:ui} highlights the five elements that comprise the \texttt{intRo} interface.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/ui_annotate.pdf}
\caption{The five elements that comprise the \texttt{intRo} application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}
 1. **Top Navigation** - The top navigation bar includes two sets of clickable icons. The left-aligned buttons are informational buttons. The first is a link to \texttt{intRo}. The second is a link to the documentation page. The third is a link to the GitHub repository where the code for \texttt{intRo} is housed. The final button is a link to our websites, which contain contact information if there are any questions or comments. The right-aligned buttons are \texttt{intRo} utilities. The first is a link to toggle the visibility of the code panel (5). The middle icon downloads an rmarkdown [@rmarkdown] document of the analysis performed. The last is a link to print the stored module results, and the associated code (if visible).
 2. **Side Navigation** - The side navigation panel includes a list of data analysis tasks.
 3. **Options Panel** - The options panel includes task-specific options which the user can use to customize their results.
 4. **Results Panel** - The results pane displays the result of the selected module and options.
 5. **Code Panel** - The code panel displays the \texttt{R} code used to generate the results from the student's \texttt{intRo} session. The code panel is shown by default to facilitate a transition to coding, but can be hidden by clicking the code toggle button in the Top Navigation bar.

The modules included in \texttt{intRo} are split into three higher level categories - data, summaries, and statistics. Under each of these categories, there are seven default modules, which perform specific data analysis tasks that employ an easy to use point-and-click interface. More modules can easily be added by an instructor, as detailed in section \ref{modules}. The default modules support uploading and downloading a dataset, transforming variables, graphical and numerical summaries, simple linear regression, contingency tables, and T-tests.

@@

\texttt{intRo} has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of \texttt{intRo}, students are actually creating a fully-executable \texttt{R} script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of \texttt{intRo} is meant to generate excitement about programming in \texttt{R} and empower students to feel that they can generate code as well. \texttt{intRo} uses \texttt{rmarkdown}'s \texttt{render} function in order to print the results, by dymanically executing the user's \texttt{R} script. By default, the output will include the \texttt{R} code, but if the user elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

On the front end, user interaction with \texttt{intRo} is split into bitesize chunks that we call modules. Modules are self-contained pieces of functionality which implement common statistical procedures that are often used in introductory statistics classes.

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from her computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of \texttt{intRo}. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in \texttt{intRo}.

@@

Each module is a self contained set of \texttt{R} code that is dynamically added to the application at run time. What this means is that \texttt{intRo} can be easily extended by the addition of modules within the framework underlying the application. This extensibility allows for \texttt{intRo} to be customized for an individual classroom setting, tailored to the needs of a particular course.

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a nonparametric framework rather than distributional. In this case, a wilcoxon rank sum test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork \texttt{intRo} from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `libraries.R`, `observe.R`, `output.R`, `reactive.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a \texttt{Shiny} application. Further details on the creation of new modules can be found in section \ref{modules}.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within \texttt{intRo}.}
\label{fig:app_creation_modules}
\end{figure*}

The process by which \texttt{intRo} is dynamically created upon running the application through module sourcing is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

  ## Source the modules
  for (mod in modules_tosource) {
      source(mod, local = TRUE)
  }
})
```
\normalsize
In this way, we were able to have \texttt{intRo} be fully extensible, its structure and functionality dependent entirely on the modules present within the application.

@@

\texttt{intRo} can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package \texttt{intRo}. Currently, the package is only available and GitHub, and can be installed using the \texttt{devtools} package as follows:

@@

After installing the \texttt{intRo} package, the first function one should call is `download_intRo`. `download_intRo` takes as an argument a directory in which to store the application. By default, it selects the working directory of the R session. This function clones the application branch of the \texttt{intRo} repository on GitHub, and hence will pull the latest version of the code whenever it is ran.

Running `download_intRo` will produce an \texttt{intRo} folder in the specified folder. It can then be ran as any \texttt{Shiny} application, using \texttt{Shiny}'s `runApp` command. However, we have provided a wrapper function `run_intRo` which adds some additional customization options to the execution process. `run_intRo` takes as argument the path to the folder containing the \texttt{intRo} application. It also takes several more optional arguments:

* `enabled_modules`: A character vector containing the modules to enable
* `theme`: A string representing a \texttt{shinythemes} theme to use
* `...`: Additional arguments passed to \texttt{Shiny}'s `runApp` function

The package provides help documentation which explains in further detail the format that these arguments would take, but as an example, suppose I wanted to download \texttt{intRo} to my working directory, execute an \texttt{intRo} session with only the data sources, data transform, and numerical summaries modules enabled, and apply the cerulean theme. The series of calls to do so would be as follows:

@@

If the intent is to use a specific instance of \texttt{intRo} where many users will access it at the same time, such as in an introductory statistics class, it may be preferable to deploy a custom instance of \texttt{intRo} to a publicly accessible URL. The package provides a function `deploy_intRo` which is a wrapper for the `deployApp` function contained in the \texttt{shinyapps} package. Once the shinyapps package is installed and configured, `deploy_intRo` will upload \texttt{intRo} as an application on the user's account. The function takes the same arguments as `run_intRo`, so it can be deployed with a custom selection of modules, and a customized theme. It also takes an additional argument `google_analytics`, which allows the specification of a Google Analytics tracking ID. It also takes `...` as additional arguments to be passed into the `deployApp` routine. For example, if we wished to deploy the instance of \texttt{intRo} we ran previously, we would call it like so:

@@

\texttt{intRo}'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as \texttt{JMP} and \texttt{Excel}, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. \texttt{JMP}, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using \texttt{R} itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like \texttt{rmarkdown}, but this again is a bit of a learning curve for the new statistics student).

We believe that \texttt{intRo} targets these key limitations. The interface is designed to be quick, seamless, and easy to understand. Because \texttt{intRo} is web-based, the statistical analysis can be performed from nearly any modern device. \texttt{R} underlies the analysis, but users need not be familiar with \texttt{R} or programming to perform it. The focal point of the interface is the output panel, allowing the student to remain focused on the results and limiting the need to fight with the interface itself to get work done. For those students who may have a natural inclination for programming, executable code is generated to allow a local reproduction of the analysis.

The features of \texttt{intRo} were selected in order to apply to the most commonly used lesson plans in early statistics courses. The textbook "The Art and Science of Learning from Data" by Agresti and Franklin [-@agresti2012statistics] references many of the focal points of introductory statistics, and \texttt{intRo} supports performing a majority of them. Part one, titled "Gathering and Exploring Data", covers graphical summaries, measures of spread and center, and avoiding the misuse of graphics in situations they are not appropriate. \texttt{intRo} includes modules for both graphical and numerical summaries covering these topics. Furthermore, \texttt{intRo} limits the display of variables to ones that are appropriate for a chosen plot type, helping to reinforce the appropriateness of particular graphics (Table \ref{tab:plottypes}).

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in \texttt{intRo}'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}

Part two of this text discusses probability and sampling distributions, which is outside the current realm of \texttt{intRo}, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in \texttt{intRo}. Finally, part four of the text discusses the association between variables, including regression, functions that \texttt{intRo} performs with ease. Some of the more advanced topics of this section are currently unsupported by \texttt{intRo}, including multiple regression.

Statistics courses with a lab component can most benefit from the use of \texttt{intRo}. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the user through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the user can easily navigate the \texttt{intRo} interface to obtain the results. For topics currently unsupported by \texttt{intRo}, the instructor can supplement its use with other web-based modules that are publicly available or create their own modules within \texttt{intRo} as detailed in section \ref{modules}.

@@

## Modules
An \texttt{intRo} module is a set of self-contained executible R scripts that together produce a set of introductory statistics functionality. \texttt{intRo} modules were designed in this way to allow for simple dynamic creation of the user interface at run-time, as well as ease the process of converting existing analysis code to the \texttt{intRo} framework. In this section, we detail the structure of these modules, and the process of creating and deploying new modules.

@@

The modules provided with \texttt{intRo} are contained in the modules folder. The top level directory in the modules folder defines the category of the module (currently `data`, `summaries`, or `statistics`). Within each of these categories is a folder named according to the name of the module. This folder houses the previously defined scripts. As an example, we will walk through the contents of the `contingency` module.

Since the `contingency` module performs a statistical analysis, it is part of the statistics category, and hence can be found in the \texttt{intRo} repository at **modules/statistics/contingency**. Let's first examine *helper.R*:

@@

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate`. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of \texttt{intRo}. More details on the reasons and choices behind `interpolate` can be found in section \ref{reproducibility}.

@@

Shiny observers are a class of reactive objects within the Shiny paradigm which do not return a value [@shiny]. For further discussion of reactivity, see section \ref{reactive-design-choices}. In this example, observers are created to ensure that the choices of variable for the `contingency` module are only categorical variables. This is accomplished by utilizing the global reactive `intro.categoricnames()`, which returns a character vector containing the variables in the current dataset that are categorical. Finally, there is an event observer to store code generated from the module into the overall code script upon clicking the store button. The presence of this observer code and the definition of the button in the user interface are enforced, and must be present in any \texttt{intRo} module.

@@

This script defines all the inputs and outputs that the user will see. The only requirements from \texttt{intRo}'s perspective are (1) that there exist a store button at the bottom of the middle panel for storing the results of the analysis in the code script, and (2) that configuration options appear in the width 4 column in the middle, and output appears in the width 8 column on the right. The remaining input and output definitions depend on the statistical analysis or transformation being performed.

Although the structure of an \texttt{intRo} module is relatively straightforward, producing the code needed in a more seamless fashion would certainly help open up the creation of such modules to a wider audience. As we discuss in the conclusions and future work section, providing an \texttt{intRo} module creation tool to abstract away some of the less common coding paradigms, like the use of `interpolate`, is an important effort that will continue to be pursued.

@@

We have designed \texttt{intRo} modules to reproduce actions taken by the user so that they can obtain \texttt{R} code associated with each module interacted with. When a user chooses to store the results associated with a particular module, R code yielding the visible results is automatically added to the code window at the bottom. This framework yields several advantages:

@@

2. An analysis created by \texttt{intRo} can be reproduced in an R session to easily assess and extend the results.

3. "Printing" the results of an \texttt{intRo} analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility.

We designed the modules in such a way that the code printed to the console is the same as the code executed by the server, which makes the process of writing an \texttt{intRo} module significantly more seamless. The workhorse function for this reproducibility framework is called `interpolate`, which is based on a function proposed by Hadley Wickham [-@interpolate]. The idea of interpolation is that by using non-standard evaluation, the call can be captured, printed, and evaluated, with all the appropriate arguments substituted into the call automatically. This means that depending on the options selected in \texttt{intRo}, the corresponding arguments to functions are appropriately set for reproduction in a standard R session.

@@

Reactive programming is a programming paradigm that ``tackles issues posed by event-driven applications by providing abstractions to express programs as reactions to external events and having the language automatically manage the flow of time (by conceptually supporting simultaneity), and data and computation dependencies.'' [@bainomugisha2012survey] As implemented by \texttt{Shiny}, results automatically update when users interact with the interface.

>>>>

The widespread adoption of R as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using R in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student. This additional time commitment may explain why introductory courses often utilize software applications such as JMP, even though instructors tend to use a tool like R in their own work. Still, some compromises must be made when using software like JMP, including dealing with software licenses and unsupported desktop platforms.

@@

Multiple software packages have recently been written in an attempt to spur student interest in introductory statistics and R programming. DataCamp's [@datacamp] courses are user-friendly ways to learning basic R programming and data analysis techniques. Swirl [@swirl] is a similar interactive tool to make learning R more fun by learning it within R itself. Project MOSAIC [@mosaic] has created a suite of tools to simplify the teaching of statistics in the form of an R package. The primary goal of DataCamp and Swirl is to teach R programming, rather than facilitate the learning of introductory statistics. Project MOSAIC's goal is to faciliate this learning, but using the package requires a knowledge of R programming that the introductory student may not have.

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text.

# What is `intRo`?

`intRo` is a web-based tool to accompany an introductory statistics class. It is meant to assist in the learning of statistics rather than as a stand-alone delivery of statistics education, with the intention of being used in conjunction with a guided class. An accompanying R package, titled `intRo` and available on GitHub, assists in the downloading, running, and deploying of `intRo` instances.

Three fundamental philosophies that guided the creation of `intRo`. In particular, `intRo` is \emph{easy} to use and can be an \emph{exciting} part of learning statistics. Additionally, `intRo` is an \emph{extensible} tool, allowing for a user of `intRo` to tailor the tool for his or her own classroom needs.

In the development of `intRo`, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in `intRo`. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{images/user_experience.pdf}
\caption{A typical user experience of generating a result in `intRo`. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.}
\label{fig:user_experience}
\end{figure}

Beyond being simple, `intRo` is also consistent. The tool is organized around specific tasks a student may perform in the process of a data analysis, called modules. To the user, a module is simply a page of statistics functionality that maintains a consistent layout, helping the user to become familiar with the location of the options, the results, and the code. Figure \ref{fig:ui} highlights the five elements that comprise the `intRo` interface.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/ui_annotate.pdf}
\caption{The five elements that comprise the `intRo` application: 1) top navigation, 2) side navigation, 3) options panel, 4) results panel, and 5) code panel.}
\label{fig:ui}
\end{figure*}
 1. **Top Navigation** - The top navigation bar includes two sets of clickable icons. The left-aligned buttons are informational buttons. The first is a link to `intRo`. The second is a link to the documentation page. The third is a link to the GitHub repository where the code for `intRo` is housed. The final button is a link to our websites, which contain contact information if there are any questions or comments. The right-aligned buttons are `intRo` utilities. The first is a link to toggle the visibility of the code panel (5). The middle icon downloads an rmarkdown [@rmarkdown] document of the analysis performed. The last is a link to print the stored module results, and the associated code (if visible).
 2. **Side Navigation** - The side navigation panel includes a list of data analysis tasks.
 3. **Options Panel** - The options panel includes task-specific options which the user can use to customize their results.
 4. **Results Panel** - The results pane displays the result of the selected module and options.
 5. **Code Panel** - The code panel displays the R code used to generate the results from the student's `intRo` session. The code panel is shown by default to facilitate a transition to coding, but can be hidden by clicking the code toggle button in the Top Navigation bar.

The modules included in `intRo` are split into three higher level categories - data, summaries, and statistics. Under each of these categories, there are seven default modules, which perform specific data analysis tasks that employ an easy to use point-and-click interface. More modules can easily be added by an instructor, as detailed in section \ref{modules}. The default modules support uploading and downloading a dataset, transforming variables, graphical and numerical summaries, simple linear regression, contingency tables, and T-tests.

@@

`intRo` has an ulterior motive as well: to get students excited about programming. By navigating about the user interface of `intRo`, students are actually creating a fully-executable R script that they can download and run locally as well as viewing the script change real-time within the application. This code creation element of `intRo` is meant to generate excitement about programming in R and empower students to feel that they can generate code as well. `intRo` uses rmarkdown's render function in order to print the results, by dymanically executing the user's R script. By default, the output will include the R code, but if the user elects to hide the source code by clicking the code toggle button at the top, the code will not appear in the printed results.

On the front end, user interaction with `intRo` is split into bitesize chunks that we call modules. Modules are self-contained pieces of functionality which implement common statistical procedures that are often used in introductory statistics classes.

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from her computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of `intRo`. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in `intRo`.

@@

Each module is a self contained set of R code that is dynamically added to the application at run time. What this means is that `intRo` can be easily extended by the addition of modules within the framework underlying the application. This extensibility allows for `intRo` to be customized for an individual classroom setting, tailored to the needs of a particular course.

As an example of possible customization, an instructor may wish to introduce hypothesis testing from a nonparametric framework rather than distributional. In this case, a wilcoxon rank sum test could be added easily as a module to be used instead of the T test. This addition would be fairly simple to complete. The user would fork `intRo` from http://github.com/gammarama/intRo and begin by adding six types of files to a folder within a module folder: `helper.R`, `libraries.R`, `observe.R`, `output.R`, `reactive.R`, and `ui.R`. Each of these files is written to comply with the reactive nature available in a Shiny application. Further details on the creation of new modules can be found in section \ref{modules}.

\begin{figure*}[ht!]
\centering
\includegraphics[width=\linewidth]{images/app_creation_modules.pdf}
\caption{This figure depicts how the \texttt{server.R} and \texttt{ui.R} files are populated using the modular structure within `intRo`.}
\label{fig:app_creation_modules}
\end{figure*}

The process by which `intRo` is dynamically created upon running the application through module sourcing is detailed in figure \ref{fig:app_creation_modules}. The key driver to populating \texttt{server.R} and \texttt{ui.R} is the navigation file \texttt{modules.txt}, which defines the placement of each module. The interface is then created with the following statement.

@@

  ## Source the modules
  for (mod in modules_tosource) {
      source(mod, local = TRUE)
  }
})
```
\normalsize
In this way, we were able to have `intRo` be fully extensible, its structure and functionality dependent entirely on the modules present within the application.

@@

`intRo` can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package `intRo`. Currently, the package is only available and GitHub, and can be installed using the devtools package as follows:

@@

After installing the `intRo` package, the first function one should call is `download_intRo`. `download_intRo` takes as an argument a directory in which to store the application. By default, it selects the working directory of the R session. This function clones the application branch of the `intRo` repository on GitHub, and hence will pull the latest version of the code whenever it is ran.

Running `download_intRo` will produce an `intRo` folder in the specified folder. It can then be ran as any Shiny application, using Shiny's `runApp` command. However, we have provided a wrapper function `run_intRo` which adds some additional customization options to the execution process. `run_intRo` takes as argument the path to the folder containing the `intRo` application. It also takes several more optional arguments:

* `enabled_modules`: A character vector containing the modules to enable
* `theme`: A string representing a shinythemes theme to use
* `...`: Additional arguments passed to Shiny's `runApp` function

The package provides help documentation which explains in further detail the format that these arguments would take, but as an example, suppose I wanted to download `intRo` to my working directory, execute an `intRo` session with only the data sources, data transform, and numerical summaries modules enabled, and apply the cerulean theme. The series of calls to do so would be as follows:

@@

If the intent is to use a specific instance of `intRo` where many users will access it at the same time, such as in an introductory statistics class, it may be preferable to deploy a custom instance of `intRo` to a publicly accessible URL. The package provides a function `deploy_intRo` which is a wrapper for the `deployApp` function contained in the shinyapps package. Once the shinyapps package is installed and configured, `deploy_intRo` will upload `intRo` as an application on the user's account. The function takes the same arguments as `run_intRo`, so it can be deployed with a custom selection of modules, and a customized theme. It also takes an additional argument `google_analytics`, which allows the specification of a Google Analytics tracking ID. It also takes `...` as additional arguments to be passed into the `deployApp` routine. For example, if we wished to deploy the instance of `intRo` we ran previously, we would call it like so:

@@

`intRo`'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as JMP and Excel, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. JMP, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using R itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like rmarkdown, but this again is a bit of a learning curve for the new statistics student).

We believe that `intRo` targets these key limitations. The interface is designed to be quick, seamless, and easy to understand. Because `intRo` is web-based, the statistical analysis can be performed from nearly any modern device. R underlies the analysis, but users need not be familiar with R or programming to perform it. The focal point of the interface is the output panel, allowing the student to remain focused on the results and limiting the need to fight with the interface itself to get work done. For those students who may have a natural inclination for programming, executable code is generated to allow a local reproduction of the analysis.

The features of `intRo` were selected in order to apply to the most commonly used lesson plans in early statistics courses. The textbook "The Art and Science of Learning from Data" by Agresti and Franklin [-@agresti2012statistics] references many of the focal points of introductory statistics, and `intRo` supports performing a majority of them. Part one, titled "Gathering and Exploring Data", covers graphical summaries, measures of spread and center, and avoiding the misuse of graphics in situations they are not appropriate. `intRo` includes modules for both graphical and numerical summaries covering these topics. Furthermore, `intRo` limits the display of variables to ones that are appropriate for a chosen plot type, helping to reinforce the appropriateness of particular graphics (Table \ref{tab:plottypes}).

\begin{table*}[ht!]
\centering
\begin{tabular}{|c|c|c|}
\hline
Plot Type&X Variable Type Allowed&Y Variable Type Allowed\\
\hline
Histogram&Numeric&N/A\\
Normal Quantile Plot&Numeric&N/A\\
Scatterplot&Numeric&Numeric\\
Line Chart&Numeric&Numeric\\
Boxplot&Categorical&Numeric\\
Bar Chart&Categorical&Numeric\\
Pareto Chart&Categorical&Numeric\\
Mosaic Plot&Categorical&Categorical\\
\hline
\end{tabular}
\caption{The plot and variable types allowed in `intRo`'s graphical module. For example, when the selected plottype is histogram, the interface displays only a single variable selection box (x), and the variables selectable in this box are only numeric. If a boxplot is selected, the x variable must be categorical and the y variable must be numeric, and the list of choices updates automatically to reflect this.}
\label{tab:plottypes}
\end{table*}

Part two of this text discusses probability and sampling distributions, which is outside the current realm of `intRo`, but part three focuses on inferential statistics. The inferential statistics include confidence intervals, hypothesis testing, and comparing two variables, all supported natively in `intRo`. Finally, part four of the text discusses the association between variables, including regression, functions that `intRo` performs with ease. Some of the more advanced topics of this section are currently unsupported by `intRo`, including multiple regression.

Statistics courses with a lab component can most benefit from the use of `intRo`. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the user through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the user can easily navigate the `intRo` interface to obtain the results. For topics currently unsupported by `intRo`, the instructor can supplement its use with other web-based modules that are publicly available or create their own modules within `intRo` as detailed in section \ref{modules}.

@@

## Modules
An `intRo` module is a set of self-contained executible R scripts that together produce a set of introductory statistics functionality. `intRo` modules were designed in this way to allow for simple dynamic creation of the user interface at run-time, as well as ease the process of converting existing analysis code to the `intRo` framework. In this section, we detail the structure of these modules, and the process of creating and deploying new modules.

@@

The modules provided with `intRo` are contained in the modules folder. The top level directory in the modules folder defines the category of the module (currently `data`, `summaries`, or `statistics`). Within each of these categories is a folder named according to the name of the module. This folder houses the previously defined scripts. As an example, we will walk through the contents of the `contingency` module.

Since the `contingency` module performs a statistical analysis, it is part of the statistics category, and hence can be found in the `intRo` repository at **modules/statistics/contingency**. Let's first examine *helper.R*:

@@

This script is most immediately similar to standard R code. In this case, a function `cont.table` is created which, depending on the values of the parameters, ultimately returns a contingency table. One important difference from a typical R script is that each call in the script is wrapped in a function called `interpolate`. `interpolate` both executes the given R code on the server, and also writes the code executed to the script window at the bottom of `intRo`. More details on the reasons and choices behind `interpolate` can be found in section \ref{reproducibility}.

@@

Shiny observers are a class of reactive objects within the Shiny paradigm which do not return a value [@shiny]. For further discussion of reactivity, see section \ref{reactive-design-choices}. In this example, observers are created to ensure that the choices of variable for the `contingency` module are only categorical variables. This is accomplished by utilizing the global reactive `intro.categoricnames()`, which returns a character vector containing the variables in the current dataset that are categorical. Finally, there is an event observer to store code generated from the module into the overall code script upon clicking the store button. The presence of this observer code and the definition of the button in the user interface are enforced, and must be present in any `intRo` module.

@@

This script defines all the inputs and outputs that the user will see. The only requirements from `intRo`'s perspective are (1) that there exist a store button at the bottom of the middle panel for storing the results of the analysis in the code script, and (2) that configuration options appear in the width 4 column in the middle, and output appears in the width 8 column on the right. The remaining input and output definitions depend on the statistical analysis or transformation being performed.

Although the structure of an `intRo` module is relatively straightforward, producing the code needed in a more seamless fashion would certainly help open up the creation of such modules to a wider audience. As we discuss in the conclusions and future work section, providing an `intRo` module creation tool to abstract away some of the less common coding paradigms, like the use of `interpolate`, is an important effort that will continue to be pursued.

@@

We have designed `intRo` modules to reproduce actions taken by the user so that they can obtain R code associated with each module interacted with. When a user chooses to store the results associated with a particular module, R code yielding the visible results is automatically added to the code window at the bottom. This framework yields several advantages:

@@

2. An analysis created by `intRo` can be reproduced in an R session to easily assess and extend the results.

3. "Printing" the results of an `intRo` analysis amounts to nothing more than executing the R code on the server, adding another layer of reproducibility.

We designed the modules in such a way that the code printed to the console is the same as the code executed by the server, which makes the process of writing an `intRo` module significantly more seamless. The workhorse function for this reproducibility framework is called `interpolate`, which is based on a function proposed by Hadley Wickham [-@interpolate]. The idea of interpolation is that by using non-standard evaluation, the call can be captured, printed, and evaluated, with all the appropriate arguments substituted into the call automatically. This means that depending on the options selected in `intRo`, the corresponding arguments to functions are appropriately set for reproduction in a standard R session.

@@

Reactive programming is a programming paradigm that ``tackles issues posed by event-driven applications by providing abstractions to express programs as reactions to external events and having the language automatically manage the flow of time (by conceptually supporting simultaneity), and data and computation dependencies.'' [@bainomugisha2012survey] As implemented by Shiny, results automatically update when users interact with the interface.

intRo leverages the reactive programming nature of Shiny, and as such is designed around the idea of user input cascading through the entire application. In a typical Shiny application, users interact with inputs that act as parameters to function, which in turn yield different results. Within `intRo`, the users are able to interact with and manipulate the data underlying the entire application. This posed many challenges in the creation of `intRo` and drove design decisions, namely timely save points according to the user's workflow, and reactive updating of variable lists tied to inputs across the entire application. Because the user may experiment with different configurations or select different variables, we did not want to store all actions taken in the intRo session. Rather, each module includes a button allowing the user to explicitly store the output visible in the results panel into the R script. This way, output is only stored when the user is satisfied, and the resulting output is not cluttered with unnecessary information.
